---
title: "R Coding Exercise"
---

# Assignment#3: Coding Basics

## Loading and checking data

```{r}
#load necessary packages
library("dslabs")
library("dplyr")
library("ggplot2")

#look at help file for gapminder data
help(gapminder)
```

```{r}
#get an overview of data structure
str(gapminder)
```

```{r}
#get a summary of data
summary(gapminder)
```

```{r}
#determine the type of object gapminder is
class(gapminder)
```

## Processing Data

```{r}
# Assigning only the African countries to a new object/variable called africadata
africadata <- gapminder %>% filter(continent == "Africa")
```

```{r}
# Display the structure of the new object
str(africadata)
```

```{r}
# Display the summary of the new object
summary(africadata)
```

### Creating two new objects:

```{r}
# Creating a new object containing only 'infant_mortality' and 'life_expectancy'
infant_life_expectancy <- africadata %>% select(infant_mortality, life_expectancy)

# Creating a new object containing only 'population' and 'life_expectancy'
population_life_expectancy <- africadata %>% select(population, life_expectancy)
```

```{r}
# Display the structure of the 'infant_life_expectancy' object
str(infant_life_expectancy)

# Display the summary of the 'infant_life_expectancy' object
summary(infant_life_expectancy)

```

```{r}
# Display the structure of the 'population_life_expectancy' object
str(population_life_expectancy)

# Display the summary of the 'population_life' object
summary(population_life_expectancy)
```

## Plotting

```{r}
# Creating the plot for life expectancy as a function of infant mortality
ggplot(data = infant_life_expectancy, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point() +
  labs(title = "Life Expectancy vs Infant Mortality",
       x = "Infant Mortality",
       y = "Life Expectancy") +
  theme_minimal()
```

```{r}
# Creating the plot for life expectancy as a function of population size with x-axis log scale
ggplot(data = population_life_expectancy, aes(x = population, y = life_expectancy)) +
  geom_point() +
  scale_x_log10() +
  labs(title = "Life Expectancy vs Population Size (Log Scale)",
       x = "Population (Log Scale)",
       y = "Life Expectancy") +
  theme_minimal()
```

## More data processing

```{r}
# Identifying years with missing data for infant mortality
missing_infant_mortality <- africadata %>%
  filter(is.na(infant_mortality)) %>%
  group_by(year) %>%
  summarise(missing_count = n())

# Display the years with missing data for infant mortality
print(missing_infant_mortality)
```

```{r}
# Extracting data for the year 2000
africadata_2000 <- africadata %>% filter(year == 2000)

# Display the structure of the new object
str(africadata_2000)
```

```{r}
# Display the summary of the new object
summary(africadata_2000)
```

## More plotting

```{r}
# Creating the plot for life expectancy as a function of infant mortality for the year 2000
ggplot(data = africadata_2000, aes(x = infant_mortality, y = life_expectancy)) +
  geom_point() +
  labs(title = "Life Expectancy vs Infant Mortality (Year 2000)",
       x = "Infant Mortality",
       y = "Life Expectancy") +
  theme_minimal()
```

```{r}
# Create the plot for life expectancy as a function of population size for the year 2000 with x-axis log scale
ggplot(data = africadata_2000, aes(x = population, y = life_expectancy)) +
  geom_point() +
  scale_x_log10() +
  labs(title = "Life Expectancy vs Population Size (Log Scale, Year 2000)",
       x = "Population (Log Scale)",
       y = "Life Expectancy") +
  theme_minimal()
```

Observations:

1.  There is a negative correlation between infant mortality and life expectancy.
2.  There is positive correlation between population size and life expectancy.
3.  The streaks of data mainly due to different years for individual countries.

## Simple fit model

```{r}
# Fit a linear model with life expectancy as the outcome and infant mortality as the predictor
fit1 <- lm(life_expectancy ~ infant_mortality, data = africadata_2000)

# Display the summary of the first fit
summary(fit1)

```

Fit 1 Summary:

The p-value is lower than the significance level of 0.05, which indicates a statistically significant negative correlation between infant mortality and life expectancy.

```{r}
# Fit a linear model with life expectancy as the outcome and population size as the predictor (log scale)
fit2 <- lm(life_expectancy ~ log(population), data = africadata_2000)

# Display the summary of the second fit
summary(fit2)
```

Fit 2 Summary:

The p-value is higher than the significance level of 0.05, which indicates taht there is no statistically significant correlation between population size (log scale) and life expectancy.
