---
title: "Assignment #5: CDC Data Exercise"
editor: visual
---

# About the Dataset from CDC's data website:

**Dataset Name:** **NCHS - Death Rates and Life Expectancy at Birth**

This dataset provides an annual data on death rates and life expectancy at birth in the United States, offering key indicator for public health analysis.

**Key variables:**

**Year:** The specific year of the data

**Race:** Black or White

**Sex:** Male or Female

**Average Life Expectancy (Years):** The average number of years a newborn is expected to live.

**Age-adjusted Death Rate:** The number of deaths per 100,000 population.

# Data Loading:

## Load necessary packages

```{r}
# Load necessary packages
library (here)
library(readr)
library(dplyr)
library(ggplot2)
```

## Read and Load the dataset

```{r}
# Define data location using here() package
data_location = here::here("cdcdata-exercise","NCHS_-_Death_rates_and_life_expectancy_at_birth.csv")

# Load Data
cdc_data = read_csv(data_location)
head(cdc_data)
str(cdc_data)
# View the column names
print(colnames(cdc_data))
```

## Renamed the columns

**The column names were too long so in order to make referencing the column name while coding I renamed two columns as shown in the code below**

```{r}
 # Rename columns to avoid issues with spaces or special characters
cdc_data = cdc_data %>%
  rename(
    DeathRate = `Age-adjusted Death Rate`,
    LifeExpectancy = `Average Life Expectancy (Years)`
  )
```

## Filter the Data

**The dataset had information from 1900 to 2018. Hence, I chose years 1993 through 2018.**

```{r}
# Filter the data
filtered_cdc_data <- cdc_data %>%
  filter(Year >= 1993 & Year <= 2018, 
         Race %in% c("Black", "White"), 
         Sex %in% c("Male", "Female"))


# DIsplay Structure of the filtered data
str(filtered_cdc_data)
```

**There were a few columns with missing values, those were removed as shown below**

```{r}
# Remove rows with missing values
cleaned_cdc_data = filtered_cdc_data %>%
  na.omit()

str(cleaned_cdc_data)

```

# Summarize cleaned data

**Summarized the cleaned data to see mean and median for the numerical variables.**

```{r}
# Let's summarize the data
summary(cleaned_cdc_data)
```

**Analysis:**

**Life Expectancy:** The distribution of life expectancy values is relatively symmetric, with the mean and median being very close.

**Death Rate:** The distribution of death rates is slightly skewed to the right, as indicated by the mean being higher than the median.

## Summarize character variables:

**First let's summarize race**

```{r}
# Summary statistics for Race
race_summary = cleaned_cdc_data %>%
  group_by(Race) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100)

print(race_summary)
```

**then let's summarize sex**

```{r}
# Summary statistics for Sex
sex_summary = cleaned_cdc_data %>%
  group_by(Sex) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100)

print(sex_summary)

```

# Visualizations

**Distribution of DeathRate**

```{r}
# Distribution of Death Rate
p1 = ggplot(cleaned_cdc_data, aes(x = DeathRate)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Death Rate", x = "Death Rate", y = "Frequency")

print(p1)
```

**Observation:** The distribution appears to be highly varied and does not follow a typical bell-curve shape. The histogram shows many individual bars, indicating a wide spread of death rates with many unique values. The death rates range from approximately 500 to over 1500. The graph indicates that the death rates in the dataset are highly variable and spread across a wide range of values.

**Distribution of Life Expectancy**

```{r}
# Distribution of Life Expectancy
p2 = ggplot(cleaned_cdc_data, aes(x = LifeExpectancy)) +
  geom_histogram(binwidth = 1, fill = "green", color = "black") +
  labs(title = "Distribution of Life Expectancy", x = "Life Expectancy", y = "Frequency")

print(p2)

```

**Observation:** The distribution of life expectancy is not perfectly normal (bell-shaped). Instead, it shows multiple peaks. Life expectancy values range from approximately 65 to 80 years. The graph shows that most of the individuals in the dataset have life expectancies around 75 and 80 years, with fewer individuals having life expectancies in the lower 70s and upper 60s.

**Race distribution**

```{r}
# Race distribution
p3 = ggplot(race_summary, aes(x = Race, y = percentage, fill = Race)) +
  geom_bar(stat = "identity") +
  labs(title = "Race Distribution", x = "Race", y = "Percentage")

print(p3)

```

**Observation:** The bar plot shows that the dataset is almost evenly split between Black and White individuals, each constituting about half of the total dataset. This balanced distribution is beneficial for analyses that compare these racial groups, as it minimizes the risk of bias due to unequal representation.

**Sex distribution**

```{r}
# Sex distribution
p4 = ggplot(sex_summary, aes(x = Sex, y = percentage, fill = Sex)) +
  geom_bar(stat = "identity") +
  labs(title = "Sex Distribution", x = "Sex", y = "Percentage")

print(p4)

```

**Observation:** The bar plot shows that the dataset is almost evenly split between females and males, each constituting about half of the total dataset. This balanced distribution is beneficial for analyses that compare these gender groups, as it minimizes the risk of bias due to unequal representation.

# **This section is contributed by Savanna Torres.**

I utilized ChatGPT to produce the synthetic data set. This is the prompt I used:

Produce a synthetic data set similar to this with both Black and White randomly for Race and Male and Female randomly for Sex:

tibble \[100 × 5\] (S3: tbl_df/tbl/data.frame)
\$ Year : num \[1:100\] 1993 1994 1995 1996 1997 ...
\$ Race : chr \[1:100\] "Black" "Black" "Black" "Black" ...
\$ Sex : chr \[1:100\] "Female" "Female" "Female" "Female" ...
\$ LifeExpectancy: num \[1:100\] 73.7 73.9 73.9 74.2 74.7 74.8 74.7 75.1 75.3 75.4 ...
\$ DeathRate : num \[1:100\] 969 954 956 940 922 ...
 - attr(\*, "na.action")= 'omit' Named int \[1:4\] 26 52 78 104
..- attr(\*, "names")= chr \[1:4\] "26" "52" "78" "104"

**Load libraries**

```{r}
# Load necessary libraries
library(tibble)
library(dplyr)
```

**Generate synthetic data:**

```{r}
# Set seed for reproducibility
set.seed(123)

# Generate data
Year <- 1993:2092  # 100 years
Race <- sample(c("Black", "White"), 100, replace = TRUE)
Sex <- sample(c("Female", "Male"), 100, replace = TRUE)
LifeExpectancy <- round(rnorm(100, mean = 75, sd = 2), 1)  # Normal distribution with mean 75 and sd 2
DeathRate <- round(rnorm(100, mean = 950, sd = 30), 0)  # Normal distribution with mean 950 and sd 30

# Combine into a tibble
synthetic_data <- tibble(
  Year = Year,
  Race = Race,
  Sex = Sex,
  LifeExpectancy = LifeExpectancy,
  DeathRate = DeathRate
)
```

**View synthetic data:**

```{r}
print(synthetic_data)
```

**Summarizing the synthetic data set:**

```{r}
summary(synthetic_data)
```

Beginning with the summary of the synthetic data we find that the minimum Life Expectancy is higher at 70.90 as opposed to the minimum Life Expectancy of 64.60 with the cleaned data set. As for the Max Life Expectancy with the synthetic data we have a slightly higher value of 81.50 while the cleaned data set's Max Life Expectancy is at 81.40.

**Summarizing character variable Sex:**

```{r}
# Summary statistics for Race
race_summary_syn = synthetic_data %>%
  group_by(Race) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100)

print(race_summary_syn)
```

The summary stats for Race with the synthetic data set are randomized here as opposed to the clean data set where it was split down the line at 50/50.

```{r}
# Summary statistics for Sex
sex_summary_syn = synthetic_data %>%
  group_by(Sex) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100)

print(sex_summary_syn)
```

Again with the summary stats for Sex using the synthetic data set we have a randomized percentage for both male and female while the cleaned data set was split down the line at 50/50.

**Distribution of Death Rate Visualization**

```{r}
# Distribution of Death Rate
p1_syn = ggplot(synthetic_data, aes(x = DeathRate)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Death Rate", x = "Death Rate", y = "Frequency")

print(p1_syn)
```

The distribution of Death Rate with the synthetic data takes more of a bell-shaped curve than the cleaned data which showed to be right skewed.

**Distribution of Life Expectancy visualization**

```{r}
# Distribution of Life Expectancy
p2_syn = ggplot(synthetic_data, aes(x = LifeExpectancy)) +
  geom_histogram(binwidth = 1, fill = "green", color = "black") +
  labs(title = "Distribution of Life Expectancy", x = "Life Expectancy", y = "Frequency")

print(p2_syn)
```

The distribution of Life Expectancy using the synthetic data is also more close to a bell curve shape while the cleaned data took more of a left skewed shape.

**Race distribution visualization**

```{r}
# Race distribution
p3_syn = ggplot(race_summary_syn, aes(x = Race, y = percentage, fill = Race)) +
  geom_bar(stat = "identity") +
  labs(title = "Race Distribution", x = "Race", y = "Percentage")

print(p3_syn)
```

We can see here visually that the synthetic data randomly distributed the percentage between race.

**Sex distribution visualization**

```{r}
# Sex distribution
p4_syn = ggplot(sex_summary_syn, aes(x = Sex, y = percentage, fill = Sex)) +
  geom_bar(stat = "identity") +
  labs(title = "Sex Distribution", x = "Sex", y = "Percentage")

print(p4_syn)
```

Also seen here is the random distribution of the percentage between sex.
